---
title: "BadranSeq JOSS Paper — Figure Generation"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

## Overview

This script generates all figures for the BadranSeq JOSS paper using the bundled `pbmc3k` dataset. Each section is self-contained and produces a single publication-ready SVG vector graphic.

Figures are saved to `../write/figures/` and referenced by `paper.qmd` via `badranseq_joss_figures/write/figures/`.

| Figure | File | Description |
|---------------------|------------------|----------------------------------|
| 1 | `fig1_comparison.svg` | UMAP default comparison (vertical) |
| 2 | `fig2_pca_variance.svg` | PCA comparison (Seurat vs BadranSeq) |
| 3 | `fig3_split_silhouette.svg` | Split-panel context preservation comparison |
| 4 | `fig4_featureplot.svg` | Feature expression comparison (2×3) |
| 5 | `fig5_elbow_colors.svg` | Elbow plot comparison (vertical) |
| 6 | `fig6_stats_violin.svg` | Statistical violin comparison (side-by-side) |
| 7 | `fig7_sankey.svg` | Sankey/alluvial diagram for metadata relationships |
| 8 | `fig8_dependency_comparison.svg` | Dependency footprint comparison (bar chart) |

## Figure 1: UMAP Default Comparison

This figure compares default UMAP output from Seurat and BadranSeq.

- **Top panel**: Seurat default output for exploratory visualisation
- **Bottom panel**: BadranSeq default output with publication-oriented styling

```{r}
#| label: fig1-comparison

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
p_seurat_default <- Seurat::DimPlot(pbmc3k, reduction = "umap") +
  ggtitle("Seurat: Default") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_badran_default <- BadranSeq::do_UmapPlot(pbmc3k) +
  ggtitle("BadranSeq: Default") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_combined <- p_seurat_default / p_badran_default +
  plot_annotation(
    title = "UMAP Default Comparison",
    subtitle = "Seurat (top) produces exploratory output; BadranSeq (bottom) renders publication-ready styling by default",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig1_comparison.svg",
  p_combined,
  width = 16,
  height = 14,
  bg = "white"
)
```

## Figure 2: PCA Comparison

This figure contrasts Seurat's default PCA visualisation with BadranSeq's automatic variance annotation.

- **Top row**: Seurat PCA plots with default axis labels
- **Bottom row**: BadranSeq PCA plots with variance explained annotated on axes

```{r}
#| label: fig2-pca-variance

# Libraries ----------

library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------

data(pbmc3k)

# Processing ----------
p_seurat_pc12 <- Seurat::DimPlot(pbmc3k, reduction = "pca", dims = c(1, 2)) +
  ggtitle("Seurat: PC1 vs PC2") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_seurat_pc23 <- Seurat::DimPlot(pbmc3k, reduction = "pca", dims = c(2, 3)) +
  ggtitle("Seurat: PC2 vs PC3") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_badran_pc12 <- BadranSeq::do_PcaPlot(pbmc3k, dims = c(1, 2)) +
  ggtitle("BadranSeq: PC1 vs PC2")

p_badran_pc23 <- BadranSeq::do_PcaPlot(pbmc3k, dims = c(2, 3)) +
  ggtitle("BadranSeq: PC2 vs PC3")

p_combined <- (p_seurat_pc12 + p_seurat_pc23) / (p_badran_pc12 + p_badran_pc23) +
  plot_annotation(
    title = "PCA Comparison",
    subtitle = "Seurat (top) omits variance explained; BadranSeq (bottom) annotates axes automatically",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------

ggsave(
  "../write/figures/fig2_pca_variance.svg",
  p_combined,
  width = 14,
  height = 10,
  bg = "white"
)
```

## Figure 3: Split-Panel Context Preservation Comparison

This figure demonstrates how Seurat's default faceting destroys spatial context while BadranSeq's silhouette approach preserves it across split panels.

- **Top row**: Seurat's `split.by` faceting — each panel shows only its subset, losing embedding geometry context
- **Bottom row**: BadranSeq's silhouette approach — grey background shows all cells, coloured overlay shows target subset, preserving spatial relationships

```{r}
#| label: fig3-split-silhouette

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
set.seed(42)
pbmc3k$condition <- sample(
  c("Control", "Treatment"),
  ncol(pbmc3k),
  replace = TRUE
)

# Seurat split (top row) - loses context
p_seurat_split <- Seurat::DimPlot(pbmc3k, reduction = "umap", split.by = "condition") +
  plot_annotation(
    title = "Seurat: split.by faceting",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# BadranSeq split (bottom row) - preserves context via silhouette
p_badran_split <- BadranSeq::do_UmapPlot(pbmc3k, split.by = "condition") +
  plot_annotation(
    title = "BadranSeq: silhouette approach",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# Combine: Seurat on top, BadranSeq below
# wrap_elements() is required here: both DimPlot(split.by=) and do_UmapPlot(split.by=)
# return patchwork objects. Nesting patchwork / patchwork causes the outer
# plot_annotation() to override inner row titles. wrap_elements() treats each
# annotated patchwork as an opaque unit, preserving its title.
p_combined <- wrap_elements(full = p_seurat_split) / wrap_elements(full = p_badran_split) +
  plot_annotation(
    title = "Split-Panel Comparison: Context Preservation",
    subtitle = "Seurat (top) isolates each group, destroying spatial context. BadranSeq (bottom) preserves full embedding geometry via grey silhouettes.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig3_split_silhouette.svg",
  p_combined,
  width = 14,
  height = 12,
  bg = "white"
)
```

## Figure 4: Feature Expression Comparison

This figure compares gene expression visualisation between Seurat's `FeaturePlot()` and BadranSeq's `do_FeaturePlot()`.

- **Top row**: Seurat's default FeaturePlot — basic colour scales, no cell borders
- **Bottom row**: BadranSeq's do_FeaturePlot — viridis scales, cell borders, publication-ready styling
- **Genes**: CD3D (T cells), CD8A (cytotoxic T cells), CD14 (monocytes)

```{r}
#| label: fig4-featureplot

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
# Seurat feature plots (top row)
p_seurat_features <- Seurat::FeaturePlot(
  pbmc3k,
  features = c("CD3D", "CD8A", "CD14"),
  reduction = "umap",
  ncol = 3
) +
  plot_annotation(
    title = "Seurat: FeaturePlot()",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# BadranSeq feature plots (bottom row)
p_badran_features <- BadranSeq::do_FeaturePlot(
  pbmc3k,
  features = c("CD3D", "CD8A", "CD14"),
  reduction = "umap",
  ncol = 3
) +
  plot_annotation(
    title = "BadranSeq: do_FeaturePlot()",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# Combine: Seurat on top, BadranSeq below
# wrap_elements() required: FeaturePlot() and do_FeaturePlot() both return patchwork
# objects for multi-feature plots. See fig3 comment for explanation.
p_combined <- wrap_elements(full = p_seurat_features) / wrap_elements(full = p_badran_features) +
  plot_annotation(
    title = "Feature Expression Visualisation Comparison",
    subtitle = "Seurat (top) uses default colour scales. BadranSeq (bottom) applies viridis scales, cell borders for legibility, and consistent publication styling.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig4_featureplot.svg",
  p_combined,
  width = 14,
  height = 10,
  bg = "white"
)
```

## Figure 5: Elbow Plot Comparison

This figure compares Seurat's `ElbowPlot()` with BadranSeq's `EnhancedElbowPlot()`.

- **Top panel**: Seurat's default elbow plot showing standard deviation by principal component
- **Bottom panel**: BadranSeq's enhanced elbow plot showing variance explained with an annotated cutoff

```{r}
#| label: fig5-elbow-colors

# Libraries ----------

library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------

data(pbmc3k)

# Processing ----------
p_seurat_elbow <- Seurat::ElbowPlot(pbmc3k, ndims = 30) +
  ggtitle("Seurat: ElbowPlot()") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_badran_elbow <- BadranSeq::EnhancedElbowPlot(pbmc3k, ndims = 30, cutoff_pc = 10) +
  ggtitle("BadranSeq: EnhancedElbowPlot()") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_combined <- p_seurat_elbow / p_badran_elbow +
  plot_annotation(
    title = "Elbow Plot Comparison",
    subtitle = "Seurat (top) shows standard deviation by principal component. BadranSeq (bottom) adds variance explained and a cutoff annotation.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------

ggsave(
  "../write/figures/fig5_elbow_colors.svg",
  p_combined,
  width = 10,
  height = 10,
  bg = "white"
)
```

## Figure 6: Statistical Violin Comparison

This figure compares distribution visualisation between Seurat's `VlnPlot()` and BadranSeq's `do_StatsViolinPlot()`, highlighting the integrated statistical testing capabilities.

- **Left**: Seurat's VlnPlot — basic violin without statistical annotation
- **Right**: BadranSeq's do_StatsViolinPlot — automatic Kruskal–Wallis test and Dunn's pairwise comparisons

```{r}
#| label: fig6-stats-violin

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
selected_annotations <- c("Naive CD4 T", "Memory CD4 T", "CD8 T")
pbmc3k_subset <- subset(pbmc3k, subset = seurat_annotations %in% selected_annotations)
pbmc3k_subset$seurat_annotations <- factor(
  pbmc3k_subset$seurat_annotations,
  levels = selected_annotations
)

# Seurat violin (left) - no statistics
p_seurat_vln <- Seurat::VlnPlot(
  pbmc3k_subset,
  features = "CD3D",
  group.by = "seurat_annotations",
  pt.size = 0
) +
  ggtitle("Seurat: VlnPlot()") +
  theme(plot.title = element_text(face = "bold", size = 12))

# BadranSeq violin (right) - with statistics
p_badran_vln <- BadranSeq::do_StatsViolinPlot(
  pbmc3k,
  features = "CD3D",
  group.by = "seurat_annotations",
  group.levels = selected_annotations
) +
  ggtitle("BadranSeq: do_StatsViolinPlot()") +
  theme(plot.title = element_text(face = "bold", size = 12))

# Combine side by side
p_combined <- p_seurat_vln + p_badran_vln +
  plot_annotation(
    title = "Distribution Comparison with Statistical Testing",
    subtitle = "Seurat (left) shows expression distributions. BadranSeq (right) integrates Kruskal–Wallis testing and Dunn's pairwise comparisons with automatic formatting.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig6_stats_violin.svg",
  p_combined,
  width = 14,
  height = 14,
  bg = "white"
)
```

## Sankey / Alluvial Diagram

This figure demonstrates `do_SankeyPlot()` showing the relationship between cluster assignments and biological cell type annotations.

- **Columns**: `seurat_clusters` → `seurat_annotations`
- **Key feature**: Each stratum is labelled and coloured; flows show how cells distribute across categories
- **Use case**: Quickly assess whether clusters map cleanly to cell types

```{r}
#| label: fig7-sankey

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)

# Inputs ----------
data(pbmc3k)

# Processing ----------
p_sankey <- BadranSeq::do_SankeyPlot(
  pbmc3k,
  columns = c("seurat_clusters", "seurat_annotations")
)

# Outputs ----------
ggsave(
  "../write/figures/fig7_sankey.svg",
  p_sankey,
  width = 9,
  height = 6,
  bg = "white"
)
```

## Figure 8: Dependency Footprint Comparison

This figure quantifies the dependency burden difference between BadranSeq and SCpubr, illustrating the trade-off between functionality and maintenance overhead.

- **BadranSeq**: 19 direct dependencies (minimal footprint)
- **SCpubr**: 56 direct dependencies (broad functionality)
- **Implication**: Reduced installation time, fewer version conflicts, lower long-term maintenance burden

```{r}
#| label: fig8-dependency-comparison

# Libraries ----------
library(ggplot2)

# Inputs ----------
# Dependency counts from DESCRIPTION files
dependency_data <- data.frame(
  Package = c("BadranSeq", "SCpubr"),
  Dependencies = c(19, 56),
  stringsAsFactors = FALSE
)

# Processing ----------
p_deps <- ggplot(
  dependency_data,
  aes(x = Dependencies, y = reorder(Package, Dependencies))
) +
  geom_col(
    aes(fill = Package),
    width = 0.6,
    show.legend = FALSE
  ) +
  geom_text(
    aes(label = Dependencies),
    hjust = -0.3,
    size = 5,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("BadranSeq" = "#2E86AB", "SCpubr" = "#A23B72")) +
  scale_x_continuous(
    limits = c(0, 70),
    breaks = seq(0, 70, by = 10),
    expand = c(0, 0)
  ) +
  labs(
    title = "Dependency Footprint Comparison",
    subtitle = "BadranSeq maintains a 66% smaller dependency tree than SCpubr (19 vs 56 direct dependencies)",
    x = "Number of Direct Dependencies",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40"),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(colour = "grey40"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )

# Outputs ----------
ggsave(
  "../write/figures/fig8_dependency_comparison.svg",
  p_deps,
  width = 10,
  height = 5,
  bg = "white"
)
```
