---
title: "BadranSeq JOSS Paper — Figure Generation"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

## Overview

This script generates all figures for the BadranSeq JOSS paper using the
bundled `pbmc3k` dataset. Each section is self-contained and produces a
single publication-ready SVG vector graphic.

Figures are saved to `../write/figures/` and referenced by `paper.qmd` via
`badranseq_joss_figures/write/figures/`.

| Figure | File | Description |
|--------|------|-------------|
| 1 | `fig1_comparison.svg` | Seurat default vs BadranSeq UMAP |
| 2 | `fig2_pca_variance.svg` | PCA with automatic variance labels |
| 3 | `fig3_split_silhouette.svg` | Split-panel silhouette visualisation |
| 4 | `fig4_featureplot.svg` | Gene expression overlay |
| 5 | `fig5_elbow_colors.svg` | Elbow plot with cutoff + colour palette |
| 6 | `fig6_stats_violin.svg` | Statistical violin plot with pairwise testing |
| 7 | `fig7_sankey.svg` | Sankey/alluvial diagram for metadata relationships |

## Figure 1: Comprehensive UMAP Comparison

This figure demonstrates the visual differences between Seurat's `DimPlot()`
and BadranSeq's `do_UmapPlot()` across default and enhanced configurations.

- **Top row**: Seurat outputs — basic exploratory plots lacking publication features
- **Bottom row**: BadranSeq outputs — publication-ready with automatic styling
- **Columns**: Default settings (left) vs enhanced configurations (right)

```{r}
#| label: fig1-comparison

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
# Seurat plots (top row)
p_seurat_default <- Seurat::DimPlot(pbmc3k, reduction = "umap") +
  ggtitle("Seurat: Default") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_seurat_labels <- Seurat::DimPlot(pbmc3k, reduction = "umap", label = TRUE) +
  ggtitle("Seurat: With Labels") +
  theme(plot.title = element_text(face = "bold", size = 12))

# BadranSeq plots (bottom row)
p_badran_default <- BadranSeq::do_UmapPlot(pbmc3k) +
  ggtitle("BadranSeq: Default") +
  theme(plot.title = element_text(face = "bold", size = 12))

p_badran_enhanced <- BadranSeq::do_UmapPlot(pbmc3k, label = TRUE, plot_cell_borders = TRUE) +
  ggtitle("BadranSeq: Enhanced") +
  theme(plot.title = element_text(face = "bold", size = 12))

# Patchwork layout: 2 rows x 2 columns
p_combined <- (p_seurat_default + p_seurat_labels) / (p_badran_default + p_badran_enhanced) +
  plot_annotation(
    title = "UMAP Visualisation Comparison",
    subtitle = "Seurat (top) requires manual styling for publication; BadranSeq (bottom) produces publication-ready output by default",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig1_comparison.svg",
  p_combined,
  width = 14,
  height = 10,
  bg = "white"
)
```

## PCA with Variance Labels

This figure showcases BadranSeq's automatic variance-explained annotation on
PCA axes — a feature absent from Seurat's default plotting.

- **Left panel**: PC1 vs PC2 with variance percentages
- **Right panel**: PC2 vs PC3 with variance percentages
- **Key feature**: Axis labels automatically formatted as "PC1 (X.X%)"
- **Rationale**: Variance explained is essential context for interpreting PCA
  plots, yet Seurat requires manual calculation and annotation

```{r}
#| label: fig2-pca-variance

# Libraries ----------

library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------

data(pbmc3k)

# Processing ----------

p_pc12 <- BadranSeq::do_PcaPlot(pbmc3k, dims = c(1, 2)) +
  ggtitle("PC1 vs PC2")

p_pc23 <- BadranSeq::do_PcaPlot(pbmc3k, dims = c(2, 3)) +
  ggtitle("PC2 vs PC3")

p_combined <- p_pc12 + p_pc23 +
  plot_annotation(
    title = "PCA with Automatic Variance Labels",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
  )

# Outputs ----------

ggsave(
  "../write/figures/fig2_pca_variance.svg",
  p_combined,
  width = 14, height = 7)
```

## Figure 3: Split-Panel Context Preservation Comparison

This figure demonstrates how Seurat's default faceting destroys spatial context
while BadranSeq's silhouette approach preserves it across split panels.

- **Top row**: Seurat's `split.by` faceting — each panel shows only its subset,
  losing embedding geometry context
- **Bottom row**: BadranSeq's silhouette approach — grey background shows all cells,
  coloured overlay shows target subset, preserving spatial relationships

```{r}
#| label: fig3-split-silhouette

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
set.seed(42)
pbmc3k$condition <- sample(
  c("Control", "Treatment"),
  ncol(pbmc3k),
  replace = TRUE
)

# Seurat split (top row) - loses context
p_seurat_split <- Seurat::DimPlot(pbmc3k, reduction = "umap", split.by = "condition") +
  ggtitle("Seurat: split.by faceting") +
  theme(plot.title = element_text(face = "bold", size = 12))

# BadranSeq split (bottom row) - preserves context via silhouette
p_badran_split <- BadranSeq::do_UmapPlot(pbmc3k, split.by = "condition") +
  plot_annotation(
    title = "BadranSeq: silhouette approach",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# Combine: Seurat on top, BadranSeq below
p_combined <- p_seurat_split / p_badran_split +
  plot_annotation(
    title = "Split-Panel Comparison: Context Preservation",
    subtitle = "Seurat (top) isolates each group, destroying spatial context. BadranSeq (bottom) preserves full embedding geometry via grey silhouettes.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig3_split_silhouette.svg",
  p_combined,
  width = 14,
  height = 12,
  bg = "white"
)
```

## Figure 4: Feature Expression Comparison

This figure compares gene expression visualisation between Seurat's
`FeaturePlot()` and BadranSeq's `do_FeaturePlot()`.

- **Top row**: Seurat's default FeaturePlot — basic colour scales, no cell borders
- **Bottom row**: BadranSeq's do_FeaturePlot — viridis scales, cell borders,
  publication-ready styling
- **Genes**: CD3D (T cells), CD8A (cytotoxic T cells), CD14 (monocytes)

```{r}
#| label: fig4-featureplot

# Libraries ----------
library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------
data(pbmc3k)

# Processing ----------
# Seurat feature plots (top row)
p_seurat_features <- Seurat::FeaturePlot(
  pbmc3k,
  features = c("CD3D", "CD8A", "CD14"),
  reduction = "umap",
  ncol = 3
) +
  plot_annotation(
    title = "Seurat: FeaturePlot()",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# BadranSeq feature plots (bottom row)
p_badran_features <- BadranSeq::do_FeaturePlot(
  pbmc3k,
  features = c("CD3D", "CD8A", "CD14"),
  reduction = "umap",
  ncol = 3
) +
  plot_annotation(
    title = "BadranSeq: do_FeaturePlot()",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )

# Combine: Seurat on top, BadranSeq below
p_combined <- p_seurat_features / p_badran_features +
  plot_annotation(
    title = "Feature Expression Visualisation Comparison",
    subtitle = "Seurat (top) uses default colour scales. BadranSeq (bottom) applies viridis scales, cell borders for legibility, and consistent publication styling.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40")
    )
  )

# Outputs ----------
ggsave(
  "../write/figures/fig4_featureplot.svg",
  p_combined,
  width = 14,
  height = 10,
  bg = "white"
)
```

## Elbow Plot and Colour Palette

This figure combines two utility outputs: the enhanced elbow plot with cutoff
annotation, and a demonstration of the BadranSeq colour palette.

- **Left panel**: `EnhancedElbowPlot()` with `cutoff_pc = 10`
  - Red dashed line marks the chosen dimensionality
  - Variance explained (%) on y-axis rather than raw standard deviation
- **Right panel**: Colour palette from `generate_badranseq_colors(10)`
  - Visualised as a bar chart showing the categorical colour scheme
  - Dark 3 qualitative palette with saturation/value adjustments

```{r}
#| label: fig5-elbow-colors

# Libraries ----------

library(BadranSeq)
library(Seurat)
library(ggplot2)
library(patchwork)

# Inputs ----------

data(pbmc3k)

# Processing ----------

p_elbow <- BadranSeq::EnhancedElbowPlot(pbmc3k, ndims = 30, cutoff_pc = 10)

colors <- BadranSeq::generate_badranseq_colors(10)
color_df <- data.frame(
  Cluster = factor(paste0("C", 0:9), levels = paste0("C", 0:9)),
  Value = 1
)

p_palette <- ggplot(color_df, aes(x = Cluster, y = Value, fill = Cluster)) +
  geom_col(width = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = setNames(colors, levels(color_df$Cluster))) +
  labs(title = "BadranSeq Colour Palette", x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

p_combined <- p_elbow + p_palette +
  plot_layout(widths = c(2, 1)) +
  plot_annotation(
    title = "Variance Diagnostics and Colour Palette",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
  )

# Outputs ----------

ggsave(
  "../write/figures/fig5_elbow_colors.svg",
  p_combined,
  width = 14, height = 5)
```

## Statistical Violin Plots

This figure demonstrates `do_StatsViolinPlot()` with automatic statistical
testing and pairwise comparisons on a subset of cell types.

- **Gene**: CD3D (T-cell marker)
- **Groups**: Naive CD4 T, Memory CD4 T, CD8 T (via `group.levels`)
- **Statistics**: Kruskal–Wallis omnibus + Dunn's pairwise (default)
- **Key feature**: Seurat-aware data extraction, automatic colour generation,
  `group.levels` for comparing specific clusters without manual subsetting

```{r}
#| label: fig6-stats-violin

# Libraries ----------

library(BadranSeq)
library(Seurat)
library(ggplot2)

# Inputs ----------

data(pbmc3k)

# Processing ----------

p_violin <- BadranSeq::do_StatsViolinPlot(
  pbmc3k,
  features = "CD3D",
  group.by = "seurat_annotations",
  group.levels = c("Naive CD4 T", "Memory CD4 T", "CD8 T")
)

# Outputs ----------

ggsave(
  "../write/figures/fig6_stats_violin.svg",
  p_violin,
  width = 8, height = 6)
```

## Sankey / Alluvial Diagram

This figure demonstrates `do_SankeyPlot()` showing the relationship between
cluster assignments and biological cell type annotations.

- **Columns**: `seurat_clusters` → `seurat_annotations`
- **Key feature**: Each stratum is labelled and coloured; flows show how cells
  distribute across categories
- **Use case**: Quickly assess whether clusters map cleanly to cell types

```{r}
#| label: fig7-sankey

# Libraries ----------

library(BadranSeq)
library(Seurat)
library(ggplot2)

# Inputs ----------

data(pbmc3k)

# Processing ----------

p_sankey <- BadranSeq::do_SankeyPlot(
  pbmc3k,
  columns = c("seurat_clusters", "seurat_annotations")
)

# Outputs ----------

ggsave(
  "../write/figures/fig7_sankey.svg",
  p_sankey,
  width = 9, height = 6)
```
